Kinematics of the Palomar 5 stellar stream from RR Lyrae stars

### 2019-08-22 paper/main.tex
disc

\section{Results} \label{sec:results}

@@

Assuming a Jacobi radius of $12^\prime$, we find 15 RRLs in the cluster tails, 11 classified as type~\typeab\ and 4 type~\typec.
Combined with the 2 type~\typeab\ and 7 type~\typec\ RRLs in the cluster itself, there are 24 RRLs in total associated to Pal~5 and its tails with membership probabilities larger than 0.5.
In total, we find the number ratio of RRL types to be $N_{ab} / N_{c} \approx 1.2$.
\figurename~\ref{fig:PA_diagram} shows the period-amplitude diagram for the RRL stars in Pal~5.
The full sample of PS1 RRLs is shown in the background for comparison, as a grayscale density histogram. The distribution of the \typeab~RRL clearly indicate Pal~5 to be an Oosterhoff I cluster, consistent with their mean period of $0.565~d$ \todo{(XX) ?}.
Our findings of the updated ratio of \rrab~to~\rrc, mean period, and Period-Amplitude distribution of the \rrab\ RRLs all support an OoI classification for Pal~5, consistent with the cluster's metallicity, $\FeH=-1.32$ \citep{TODO}.
Out of our newly found \rrab\ RRLs, three are High Amplitude Short Period (HASP, $P <0.48$~d, $A_V>0.75$~mag $=A_g>$) stars.
This type of RRL is only found in relatively metal-rich globular clusters  \citep[$\FeH>-1.5$;][]{Monelli:2017} and the Sagittarius dwarf galaxy, but not in any other dwarf spheroidal galaxy. \todo{CM:( any other point to be made about this?)}.

@@

\section{Discussion}

The Pal~5 cluster has 5 previous known RRL stars (V1-V5), all of type \typec, as reported in the \citet{Clement2001} compilation of variable stars in Galactic globular clusters and dating back to \citet{SawyerHogg1973}. Star V2 has a  systemic radial velocity of  $-65\pm16$~\kms~measured by \citet{Vivas2005}.

In their analysis of the halo density profile \citet{Vivas2006} report an overdensity, dubbed Group~6, containing 6 previously unidentified RRLs. Based on the distance to the RRLs they claim many of these could be associated to Pal~5's tidal tails. The QUEST IDs of the stars in their Group~6 are: 403, 405, 407, 411, 416 and 422. The first two stars --403 and 405-- are located at distances of  $9\farcm8$ and $11'$ from the cluster's center, which \citet{Vivas2004} notes places them within the cluster's tidal radius, assuming the value of $r_t\sim16'$ reported by \citet{Odenkirchen:2002}.  \todo{Update this value}. \citet{Vivas2006} also note that star 393, not in Group~6, could also be associated to the cluster's tails, given it's distance and location along the known extent of the tails.

In summary,  previous studies report 7 known RRLs associated to the cluster itself (1 \typeab~and 6 \typec) plus 5 more reported as possibly associated to the tails (4 \typeab~and 2 \typec).

\todo{update this with a comparison to Ibata+2017?}

@@

\todo{what did we do}

>>>>

\section{Results and Discussion} \label{sec:results}

@@

Assuming a Jacobi radius of $12^\prime$ (computed assuming $\textrm{M}_{\rm Pal 5} \approx 1.4\times10^4~\msun$ and using the Milky Way mass model in \citealt{gala}), we find 15 RRLs in the cluster tails, 11 classified as type~\typeab\ and 4 type~\typec.
Combined with the 2 type~\typeab\ and 7 type~\typec\ RRLs in the cluster itself, there are 24 RRLs in total associated to Pal~5 and its tails with membership probabilities larger than 0.5.
In total, we find the number ratio of RRL types to be $N_{ab} / N_{c} \approx 1.2$.
\figurename~\ref{fig:PA_diagram} shows the period-amplitude diagram for the RRL stars in Pal~5.
The full sample of PS1 RRLs is shown in the background for comparison, as a grayscale density histogram. The distribution of the \typeab~RRL clearly indicate Pal~5 to be an Oosterhoff I cluster, consistent with their mean period of $0.565~d$ \todo{(XX) ?}.
Our findings of the updated ratio of \rrab~to~\rrc, mean period, and Period-Amplitude distribution of the \rrab\ RRLs all support an OoI classification for Pal~5, consistent with the cluster's metallicity, $\FeH=-1.32$ \citep{TODO}.
Out of our newly found \rrab\ RRLs, three are High Amplitude Short Period (HASP, $P <0.48$~d, $A_V>0.75$~mag $=A_g>$) stars.
This type of RRL is only found in relatively metal-rich globular clusters  \citep[$\FeH>-1.5$;][]{Monelli:2017} and the Sagittarius dwarf galaxy, but not in any other dwarf spheroidal galaxy. \todo{CM:( any other point to be made about this?)}.

@@

\subsection{Previously known Pal 5 RR Lyrae}

The Pal~5 cluster has 5 previous known RRL stars (named V1-V5), all of type \typec, as reported in the \citet{Clement:2001} compilation of variable stars in Galactic globular clusters and dating back to \citet{SawyerHogg:1973}.
In their analysis of the halo density profile with RRLs, \citet{Vivas:2006} report an overdensity (``Group~6'') that they claim could be associated with Pal~5's tidal tails.
% The QUEST IDs of the stars in their Group~6 are: 403, 405, 407, 411, 416 and 422.
% The first two stars --403 and 405-- are located at distances of  $9\farcm8$ and $11'$ from the cluster's center, which \citet{Vivas2004} notes places them within the cluster's tidal radius, assuming the value of $r_t\sim16'$ reported by \citet{Odenkirchen:2002}.
Two of these stars (IDs 403 and 405) are separated by just $\sim10'$ and $11'$ from the cluster's center, which places them within the cluster's Jacobi radius \citep{Vivas:2004}, and an additional star (star 393) outside of the Jacobi radius but at a similar distance could be associated with the stream \citep{Vivas:2006}.
More recently, \citet{Ibata:2017} use a prior iteration of the PS1 RRL catalog \citep{Hernitschek:2016} to identify \todo{XX} candidate members of the Pal 5 cluster and tails based on sky position and distance alone.
However, without a published catalog, we cannot compare our sample to theirs.

% In summary,  previous studies report 7 known RRLs associated to the cluster itself (1 \typeab~and 6 \typec) plus 5 more reported as possibly associated to the tails (4 \typeab~and 2 \typec).

@@

Using a catalog of RR Lyrae-type stars identified using multi-epoch, multi-band photometry from the Pan-STARRS PS1 survey and cross-matched to \Gaia\ \DR{2}, we XX.
